---
format: html
editor: visual
---

We selected `r length(unique(detect_data$NWFSCsampleID))` water samples collected at `r length(unique(detect_data$station))` unique locations during the 2019 NOAA hake abundance survey (@fig-bubble), targeting 0, 50, 150, 300, and 500 m sampling depths at each station. Where the bottom was less than 500 m deep, we did not collect a sample at 500 m. We amplified and sequenced these samples in 1 technical replicate using MFU, 3 technical replicates using MV1, and 3-5 technical replicates using DL, resulting in a final total of `r detect_data %>% group_by(primer, NWFSCsampleID) %>% n_groups()` amplified and sequenced samples.

Across all samples, we observed `r sum(detect_data$Detected)` positive detections of `r length(unique(detect_data$BestTaxon))` cetacean species, including 6 baleen whale species, 5 dolphin species, 2 porpoise species, and 3 beaked whale species (Table 1 will include species-specific common names, abbreviations, n detections, etc). XX species were observed most frequently, and XXX least frequently. *More here about detection frequency in the overall study.* @fig-bubble illustrates the distribution of detections across sampling depths, ignoring x-y distribution, for all cetacean species included in the study.

```{r}
#| label: fig-bubble
#| fig-cap: "Left: Distribution of detections across sampling depths for all species included in the study. Species are grouped as follows: Baleen whales includes the Balaenopteridae, Balaenidae, and Eschericthidae; Dolphins/Porpoises includes all Delphinidae and Phocoenidae; Beaked whales includes all Ziphiidae. Right: Geographic distribution of samples used in this study, acquired from water collected during a 2019 survey of hake distribution in the California Current. "
#| fig-width: 8
#| fig-height: 6

aligned <- align_plots(detectDepth_bubble, map, align = "v", axis = "tblr")
plot_grid(plotlist = aligned, ncol = 2, rel_widths = c(3, 1))
```

### Question 1: Does POD vary with depth?

GAM-based model selection of detection probability across sampling depths negated the null hypothesis of no variation across depth. The best alternative model allows probability of detection to vary by time spent at depth with a species-specific smooth (AIC = 4061). *Model diagnostics here, include table of AIC values?*

Species-specific detections were broadly highest at depths where they are predicted to spend the most time, especially within the Dephinidae, with notable exceptions among the beaked whales and baleen whales (@fig-H1POD-species). We present predictions of POD by depth in @fig-H1POD-species alongside trends in relative time spent at each depth to illustrate the depth at which each species in the study is most likely to be detected, as well as which species can be detected where they spend time and which are more likely to be detected elsewhere. As noted above, all three beaked whale observed in the study are more likely to be detected at depths greater than 300m, despite being predicted to spend the majority of their time above 100m. Similarly, baleen whales rarely travel below 100m (*check this number*) but in this study two patterns emerged: two species had highest POD at the surface and 500m (fin and humpback whales), and three species had increasing POD with increasing depth (minke, blue, and bowhead whales).

```{r}
#| label: fig-H1POD-species
#| fig-cap: "Species-specific probability of detection across all depths sampled in this study. Solid line shows POD; shaded region shows 95% confidence interval. Dashed line shows proportional time at depth. Species abbreviation codes can be found in Table 1; plots are colored by taxonomic category as outlined in Table 1. "
#| fig-width: 10
#| fig-height: 6

m1.2POD

```

Our models stratifying detections by taxonomic Family and preferred prey type both had lower AIC values than the model stratified by species (AIC = `r round(filter(modelAIC, model == "m1.2a")$AIC, digits = 0)` and `r round(filter(modelAIC, model == "m1.2b")$AIC, digits = 0)`, respectively); however, these models still had significant terms for Family and depthxFamily interactions (taxonomic model) and for depth, prey type, and depthxprey type interactions (prey type model). *Model diagnostics here.* Predicted POD curves for each model illustrate varying trends in POD with depth across strata (@fig-H1POD-family-prey).

*In the discussion, talk about how ecological strata may be used to guide sampling for lesser known-species - e.g. if we don't have any information about a particular baleen whale species, we might let the prey distribution model guide our sampling strategy and choose to sample at 50-100m and \~500m (where poop is).*

```{r}
#| label: fig-H1POD-family-prey
#| fig-cap: "Left: POD by depth for species stratified by taxonomic Family. Right: POD by depth for species stratified based on broad prey categories."
#| fig-width: 10
#| fig-height: 6

m1.2aPOD + m1.2bPOD + plot_layout(widths = c(3,1), axis_titles = "collect") & 
  theme(legend.position = "bottom") 

```

#### Question 2: Does depth variability in POD covary with observation bias?

Technical replication increases probability of detection.

#### Question 3: Does depth variability in POD covary with xy variability?

*Depth does covary with xy. Don't go so far as to make 3D maps here (those will go in the Big Product, but maybe choose 1-2 species with the most samples/highest significant and 1) make figures showing depth variability at 3 latitudes holding longitude constant, and 2) figures showing depth variability at 3 longitudes holding latitude constant.*

```{r}
#| label: fig-H3PODlat
#| fig-cap: "Depth of maximum probability of detection (POD; top) and width of 95% confidence itervals at max POD depth (bottom) for three species: Bbai = Baird's beaked whale (*B. bairdii*), Lobl = Pacific white-sided dolphin (*L. obliquidens*), Mnov = humpback whale (*M. novaegliae*)."
#| fig-width: 7
#| fig-height: 10

depth_max_detect / POD_depth_CI + plot_layout(guides = "collect")

```
